Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration

### 2023-10-08 manuscript/main.tex
First proof read and edits

    \citet{kim1998stochastic} proposed a sampling strategy for stochastic volatility through a combination of conjugate posterior distributions, a Kalman filter, and Metropolis Hastings over 20 years ago. Since then, there have been many advances in statistical computing to estimate complex, high dimensional Bayesian models. In particular, the recent availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, is an advancement of MCMC for efficient sampling of complex probabilistic models. The adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as integrated nested Laplace approximation \citep{rue2009approximate}.

    These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in our statistical workfow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research conducts a simulation study to assess the algorithms used to estimate stochastic volatility models. The first algorithm replicates Kim Sherphard and Chib's (KSC) Gaussian mixture approximation which is applied on a transformed stochastic volatility model with a log chi squared error distribution. Their sampling method uses the aforementioned Kalman Filter estimating the latent states, conjugate posterior distributions and the Metropolis Hastings algorithm. The second algorithm is the Hamiltonian Monte Carlo (HMC) algorithm as implemented in the Stan programming language. The objective is to analyse the efficiency of the sampling approach and their ability to return calibrated posterior estimates.

    Comparisons of these algorithms are conducted by Simulation Based Calibration \citep{talts2018validating}. Parameters are drawn from the prior distribution and used to create datasets from the generative (stochastic volatility) model which are sampled using a MCMC method. Repeating this process multiple times gives insight to how well the sampling approach can estimate the true parameters governing the data generating process on average. The key diagnostic metrics are the effective sample size to measure the efficiency of the algorithm and visual tests of uniform rank statistics to assess calibration of the posterior estimates.

    Additionally, two different parametiersations of the stochastic volatility model are considered for each sampling method. These are described as either 'centred' or 'non-centred' in location. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC efficiency as described in \citet{strickland2008parameterisation}. The key contribution of this research is to use simulations to assess whether a MCMC algorithm can return calibrated posterior estimates as well as the efficiency under the different parametiersations.

    This paper is structured as follows. Section 2 provides the context around this research, namely the stochastic volatility model examined and the limitations of single simulation studies and MCMC convergence diagnostics. Section 3 describes the simulation design and diagnostic metrics. Section 4 details the sampling approaches. Section 5 discusses the results and limitations. Finally, section 6 concludes and provides points for further reserach.

@@

\subsection{Diagnostic limitations on real data}
    The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics are often used to check the performance of the MCMC chains, such as the effective sample size and the $\hat{R}$. These diagnostics provide evidence on whether an MCMC has failed to converge onto the target posterior distribution.

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the sampling algorithm or both.

    Furthermore, different sampling strategies and MCMC tools could provide different posterior estimates for the same model. Assuming no issues in computation, these diagnostic measures do not provide any evidence to which estimate is closer to the truth.

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned log returns of the 2023 January-September S\&P 500 Index. Figure \ref{fig:realdataex} visualises the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ with corresponding summary statistics in Tables \ref{tab:realdata}.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from the Hamiltonian Monte Carlo (HMC) and Kim Sherphard and Chib (KSC) MCMC samplers using S\&P 500 data. While the shape of the distributions are similar, the modes of method 2 is much higher and the tails of method 1 are fatter for $\phi$ and $\sigma^2$. Convergence diagnostics do not tell us which estimate is ``more correct'' since we don't observe the true model or true data generating process.}
        \label{fig:realdataex}
    \end{figure}

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{Summary statistics for KSC sampler}
        \label{tab:realdata}
    \end{table}

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$  has a fatter left tail for the KSC sampler and $\sigma^2$ has heavier right tails for HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, sampling method 2 has higher peaks at their mode relative to method 1 with a tighter spread. Based on this information alone, there is no certainty in whether method 1 or method 2 is has a more correct estimate conditional on data and model. There are other methods for model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

\subsection{Diagnostic limitations on a single simulation}
    Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Furthermore, as discussed in \citet{gelman2020bayesian}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

    Simulations enable us to check whether our models and algorithms can estimate the true data generating process, however, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{talts2018validating} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude ``that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed''.

    Figure 2 show the posterior distributions of a single simulation where data is generated and sampled from the stochastic volatility model with known parameters.

@@

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple true parameters are required. This methodology is discussed in the next section.

    Additionally, two sampling strategies will be assessed using the proposed simulation design as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Not only will the calibration of any one algorithm be assessed, but also compared across sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

        To illustrate this procedure, let $\theta$ be a parameter and $y$ represent the dataset. Start with a single draw from the prior distribution:

@@

        Then take draws from the posterior distribution generated by a MCMC algorithm or estimation strategy (Hamiltonian Monte Carlo or KSC) conditional on this dataset.

@@

        That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior samples from the prior distribution suggests that the sampling methodology is not producing the correct posteriors.

        If the posterior samples follows the prior distribution, the rank statistic for a given parameter follows a discrete uniform distribution\footnote{Proof of this result in \citet{talts2018validating}.}. The rank statistic is defined as:

@@

        This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if any problematic features exist.

        If the computation is well calibrated and the rank statistics follow a discrete uniform distribution, then the posterior credible intervals have sufficient coverage. That is, one way to describe calibration is: for any percentage interval selected over the posterior samples (for example 90\%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval. Another way of saying this is a Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations.

    \subsection{Evaluation Metrics}
        The key metrics and diagnostics to compare the performance of these methods are rank statistics, and chi-squared test statistics to measure calibration and the effective sample size (ESS) to measure effciency.

        \subsubsection{Rank statistics}
            As discussed in the SBC section, rank statistics are used to evaluate the calibration of the MCMC algorithm. If a posterior is well calibrated then it is expected that the distribution of rank statistics is uniform.

            The shape of the rank statistic distribution gives insight into how a MCMC may be miscalibrated. Specifically, it gives information about the type of bias in the posterior estimates for a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2018validating} for conveience which gives some examples of the posterior estimates which result in bias of the rank statistics.

@@

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. The sampler is disproportionately over and under estimating the true parameter. \citet{talts2018validating} describe this bias as underdispersion of the posterior relative to the prior distribuion. The estimated posterior is too narrow relative to the spread of the prior resulting in bias on both ends of the rank statistics distribution.

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/rhs.png}
                \caption{Left: Non uniform rank statistics with peak on the right side. Right: Posterior distribution vertical line representing true parameter. Bias in rank statistics distribution comes from disproportionate number of SBC iterations underestimating the true parameter.}
                \label{fig:underestimation}
            \end{figure}

@@

            \subsubsection{Chi squared statistics}
            A drawback from visualising the rank statistic distributions is there are more parameters (1003 parameters) than data points in this model. An alternative to visually checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each histogram bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$. Then the chi squared statistic is given by:

@@

            A perfectly discrete uniform distribution will return a $\chi^2$ statistic of zero. That is, the number of rank statistics in each bin is equal to the expected count under a discrete uniform distribution conditional on the number of bins and observations. Instead of performing multiple hypothesis tests for each parameter, the distribution of $\chi^2$ statistics is compared across all parameters and simulations. This will give a high level summary of calibration and overall performance of the algorithms.

@@

            The ESS is defined as $n_{eff}$ and is a function of the number of chains $M$, the number of samples per chain $N$ and the autocorrelation at lag t $\rho_t$. $T$ is a trunction lag chosen such that the sum of any two consecutive autocorrelation lags is negative. This truncation (or some kind of downweighting) is required because a large number of lags leads to noiser autocorrelation estimates. More details about this and the calculation of the autocorrelation can be found in \citep{vehtari2021rank} and \citet{geyer1992practical}.

@@

    \subsection{Sampling method 1}
        (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable. So a more recent simulation smoother is used which is written by the same author of the original software.} \citep{dejong1995}.

@@

        Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and variance 4.93. The relationship between $y_t$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution. This is defined by:

@@

        \subsubsection*{Implementation}
        The code used to replicate the KSC algorithm was written by \citet{chad2018} and ammended for the purposes of this research. Additional code was developed to produce rank statistics, quantities of interest as well as integrated into the SBC framework.

@@

@@

    \subsection{Sampling method 2}
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

        The key innovation of Hamiltonian Monte Carlo (HMC) is using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. HMC is able to reach more distant points with higher acceptance probabilites since proposals are made with information about the target distribution. Random walk beahviour is more inefficient in higher dimensions since it is more difficult to randomly select a point with a higher joint density and therefore higher acceptance probability. The increasing complexity of high dimensions make it harder for a random walk sampler to explore regions over long distances, resulting in the sampler to become stuck and requiring a lot of iterations to converge onto the target distribution.

        A thorough conceptual and theoretical explanation of HMC can be found in \citet{gelman2013bayesian} and \citet{betancourt2017conceptual} with further details found in the aforementioned literature. In summary, HMC builds upon the Metropolis Hastings algorithm through the introduction of Hamiltonian equations, the momentum variable $\phi$ and the gradients of the target log posterior distribtuion. The vector of model parameters $\theta$ is jointly sampled with a vector of momentum variables $\phi$. The momentum variables is also called auxiliary variables since they are required as part of the Hamiltonian equations to generate an efficient proposal but are not a quantity of interest for inference.

        The sampled momentum variable sets out the proposal path for $\theta$ governed by the Hamiltonian dynamics (represented by a set of differential equations). These differential equations are solved using a discrete approximation algorithm, such as a leapfrog integrator, and is a function of the log posterior gradients for a given point $\theta$. The leapfrog integrator has a set of tuning parameters which determines the size and number of steps with the final step being the proposal value $(\theta^{\ast}, \phi^{\ast})$. Finally, the Metropolis Hastings accept/reject step is applied using the parameters at the beginning of the leapfrog process $(\theta^{t-1}, \phi^{t-1})$ and the proposal values $(\theta^{\ast}, \phi^{\ast})$.

        \subsubsection{Implementation}
        The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler \citep{hoffman2014no}, allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will also use the same priors as specified in the Gaussian mixture approximation. A general description of the Hamltonian Monte Carlo algorithm is given below.

        Let $\theta$ be a target parameter and $\phi$ be the auxiliary momemtum variable. The Hamiltonian is defined as:

@@

        HMC on the other hand is a more general approach to applying MCMC. The implementation of HMC is using the Stan programming language which allows for the specfication other generative or Bayesian models, not just stochastic volatility. The implementation details of HMC is abstracted away from the user, who only needs to focus on specifying the model using the Stan syntax. Any hyperparameter tuning and initialisation is automated away. Furthermore, Stan and HMC allows for the specification of different priors, whereas KSC's bespoke algorithm is dependent on the use of conjugate priors.

@@

        The stochastic volatility model as described in section 2.1 is follows a centered parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised in a way that improves MCMC performance. Reparameterisation applies a transformation to the model which may change the interpretation of the parameters or the way they are sampled but remains mathematically equivalent to the original model. The reparametiersations available for a model is dependent on the MCMC approach. Some reparamterisations are available only to specific sampling strategies and may improve or impair a MCMC performance.

        SBC is applied to reparameteirsied versions of the stochastic volatility model available to see if there is any improvement in calibration and efficiency.

        \subsubsection{Reparameterised stochastic volatility for Hamiltonian Monte Carlo}
        The reparameterisation for Hamiltonian Monte Carlo changes how the latent states are sampled. Specifically, the states are first sampled from a standard normal distribution and are transformed to be samples from the state equation. As noted above, the centered parametersation samples the states directly on a non zero mean.

@@

        Generate a the second state parameter implied by the joint prior.

@@

\section{Appendix C: Importance weighted rank statistics for reparameterised model}

>>>>

    \citet{kim1998stochastic} proposed a sampling strategy for stochastic volatility through a combination of conjugate posterior distributions, a Kalman filter, and the Metropolis Hastings algorithm. Since then, there have been many advances in statistical computing and algorithm design. In particular, Hamiltonian Monte Carlo - a variant of the Metropolis Hastings algorithm - is an advancement in MCMC for efficient sampling of complex, high dimensional probabilistic models. The adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as integrated nested Laplace approximation \citep{rue2009approximate}.

    These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in statistical workfow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research conducts a simulation study to analyse the calibration and efficiency of MCMC used to estimate stochastic volatility models. The first algorithm replicates Kim Sherphard and Chib's (KSC) Gaussian mixture approximation which is sampled using the aforementioned Kalman Filter and Metropolis Hastings algorithm. The second algorithm is Hamiltonian Monte Carlo (HMC) as implemented in the Stan programming language. The objective is to analyse the efficiency of the MCMC and their ability to return correct posterior estimates.

    Comparisons of these algorithms are conducted by Simulation Based Calibration \citep{talts2018validating}. Parameters are drawn from the prior distribution and used to create datasets from the generative stochastic volatility model which are sampled using one of the mentioned MCMC methods. Repeating this process multiple times gives insight to how well the MCMC algorithm can estimate the true parameters governing the data generating process. The key diagnostic metrics are the effective sample size to measure the efficiency of the algorithm and visual assessment of uniform rank statistics to determine calibration of the posterior estimates.

    Additionally, two different parametiersations of the stochastic volatility model are considered for each sampling method. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC performance as described in \citet{strickland2008parameterisation}. The key contribution of this research is to use simulations to assess whether a MCMC algorithm can return calibrated posterior estimates as well as the efficiency under the different parametiersations.

    This paper is structured as follows. Section 2 provides the context around this research, namely the stochastic volatility model and the limitations of single simulation studies and MCMC convergence diagnostics. Section 3 describes the simulation design and diagnostic metrics. Section 4 details the sampling approaches. Section 5 discusses the results and limitations. Finally, section 6 concludes and provides points for further reserach.

@@

\subsection{Diagnostic limitations on real data}
    The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics are often used to check the performance of the MCMC chains, such as the effective sample size and the $\hat{R}$. These diagnostics provide evidence on whether a MCMC has failed to converge onto the target posterior distribution.

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the MCMC algorithm or both. Furthermore, different sampling strategies could provide different posterior estimates for the same model. Assuming no issues in computation, convergence diagnostics do not provide any information to which estimate is more correct or closer to the true parameter.

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned daily log returns of the 2023 January-September S\&P 500 Index. Figure \ref{fig:realdataex} visualises the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ with corresponding summary statistics in Table \ref{tab:realdata}.

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ has a fatter left tail for the KSC sampler and $\sigma^2$ is also heavier in the right tails relative to HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, HMC has higher peaks at their mode relative to KSC with a tighter spread. Based on this information alone, there is no certainty in whether method 1 or method 2 is has a more correct estimate conditional on data and model.

    % Other methods for model selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from the Hamiltonian Monte Carlo (HMC) and Kim Sherphard and Chib (KSC) MCMC samplers using S\&P 500 data. While the shape of the distributions are similar, the modes of HMC is much higher and the tails of KSC are fatter for $\phi$ and $\sigma^2$. Convergence diagnostics do not tell us which estimate is more correct since we don't observe the true model or true data generating process.}
        \label{fig:realdataex}
    \end{figure}

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{Summary statistics for HMC and KSC algorithms. KSC's 25th quantile $\phi$ is smaller than HMC which is consistent with the fatter left tail. Similarly, the KSC's 75th qunatile for $\sigma^2$ is larger consistent with a larger right tail spread. Both sets of estimate are reasonable but it is unclear from estimates on real data which estimate is closer the the truth.}
        \label{tab:realdata}
    \end{table}

\subsection{Diagnostic limitations on a single simulation}
    Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from a generative model using known parameters. Then fit the same model on the simulated data and see if the true parameters can be recovered. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Furthermore, as discussed in \citet{gelman2020bayesian}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in stochastic volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

    Simulations enable us to check whether our models and algorithms can estimate the true data generating process. However, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{talts2018validating} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude ``that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed''.

    Figure 2 show the posterior distributions of a single simulation where data is generated and posteriors are sampled from the stochastic volatility model with known parameters.

@@

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to estimate stochastic volatility models. As discussed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple true parameters are required. This methodology is discussed in the next section.

    Additionally, both the HMC and KSC algorithms will be assessed using the proposed simulation design as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Results from the study will also be used to compare both sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

        To illustrate this procedure, let $\theta$ be an arbitrary parameter from a model and $y$ represent a dataset. Start with a single draw from the prior distribution:

@@

        Then take draws from the posterior distribution generated by a MCMC algorithm (HMC or KSC) conditional on this dataset.

@@

        That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior distribution from the prior distribution means that the sampling methodology is not producing the correct posteriors.

        \citet{talts2018validating} prove that the rank statistics for a given parameter follows a discrete uniform distribution if the posterior samples follow the same distribution as the prior. The rank statistic is defined as:

@@

        This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if the algorithm is calibrated and returning the correct posteriors.

        Posterior credible intervals are said to have sufficient coverage if Bayesian computation is well calibrated and the rank statistics follow a discrete uniform distribution. That said, there are two ways to interpret calibration.

        \begin{enumerate}
            \item For any percentage interval selected over the posterior samples (for example 90\%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval.

            \item Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations. 
        \end{enumerate}

        % That is, one way to describe calibration is: for any percentage interval selected over the posterior samples (for example 90\%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval. Another way of saying this is a Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations.

    \subsection{Evaluation Metrics}
        The key metrics to compare the performance of these methods are rank statistics and chi-squared test statistics to measure calibration, and the effective sample size (ESS) to measure effciency.

        \subsubsection{Rank statistics}
            Rank statistics are used to evaluate the calibration of the MCMC algorithm. If a posterior is well calibrated then it is expected that the distribution of rank statistics is uniform.

            The shape of the rank statistic distribution gives insight into how a MCMC may be miscalibrated. Specifically, it gives information about the type of bias in the posterior estimates for a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2018validating} for convenience which display examples of posterior estimates that give biased rank statistics.

@@

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. The sampler is disproportionately over and under estimating the true parameter. \citet{talts2018validating} describe this bias as underdispersion of the posterior relative to the prior distribuion. The estimated posterior is too narrow relative to the spread of the prior resulting in bias on both ends of the rank statistic distribution.

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/rhs.png}
                \caption{Left: Non uniform rank statistics with peak on the right side. Right: Posterior distribution is biased to the left side of the prior distribution. Bias in rank statistics distribution comes from disproportionate number of SBC iterations underestimating the true parameter. Note that the two plots are independently produced for illustration only.}
                \label{fig:underestimation}
            \end{figure}

@@

            \subsubsection{Chi squared statistics}
            A drawback from visualising the rank statistic distributions is there are more parameters and latent states to estimate than data points in this model. An alternative to visually checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each histogram bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$ (where the expected count is a function of the number of bins and data points under a discrete uniform distribution). Then the chi squared statistic is given by:

@@

            A perfectly discrete uniform distribution will return a chi squared statistic of zero. That is, the number of rank statistics in each bin is equal to the expected count under a discrete uniform distribution. The distribution of chi squared statistics is visualised to compare results across multiple simulations. This will give a high level summary of calibration and overall performance of the algorithms.

@@

            The ESS is defined as $n_{eff}$ and is a function of the number of chains $M$, the number of samples per chain $N$ and the autocorrelation at lag t $\rho_t$. $T$ is a trunction lag chosen such that the sum of any two consecutive autocorrelation lags is negative. This truncation (or some kind of downweighting) is required because a large number of lags leads to noisier autocorrelation estimates. More details about this ESS formula and the calculation of the autocorrelation can be found in \citep{vehtari2021rank} and \citet{geyer1992practical}.

@@

    \subsection{Sampling method 1}
        KSC sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable. So a more recent simulation smoother is used which is written by the same author of the original software.} \citep{dejong1995}.

@@

        Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and variance 4.93. The relationship between $y_t$ and $h_t$ is now linear; however, the error is not Gaussian. Since it is not simple to sample from this parameterisation of the model, KSC use a mixture of Gaussians to approximate the first 4 moments of the log chi squared distribution. This is defined by:

@@

        \subsubsection*{Implementation}
        The code used to replicate the KSC algorithm was written by \citet{chad2018} and ammended for the purposes of this research. Additional code was developed to produce rank statistics, quantities of interest as well as to integrate the MCMC into the SBC framework.

@@

        For sampling the mixture model the Gaussian mixture density is rewritten with respect to a indicator variable $s_t$

        $s_t$ is then sambled from probability mass function:

@@

    \subsection{Sampling method 2}
        Hamiltonian Monte Carlo (HMC) is a MCMC algorithm which has become widely available for efficiently sampling sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

        The key innovation of HMC is using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. HMC is able to reach more distant points with higher acceptance probabilites since proposals are made with information about the target distribution. Random walk samplers are become inefficient in higher dimensions since it is more difficult to randomly select a point with a high acceptance probability. The increasing complexity of high dimensions make it harder for a random walk sampler to explore regions over long distances, resulting in the sampler to become stuck and requiring a lot of iterations to converge onto the target distribution.

        A thorough conceptual and theoretical explanation of HMC can be found in \citet{gelman2013bayesian} and \citet{betancourt2017conceptual} with further details found in the aforementioned literature. HMC builds upon the Metropolis Hastings algorithm through the introduction of Hamiltonian equations, auxiliary momentum variables $\phi$, and the gradients of the target log posterior distribtuion. The vector of model parameters $\theta$ is jointly sampled with a vector of momentum variables $\phi$. The momentum variables are also called auxiliary variables since they are required as part of the Hamiltonian equations to generate an efficient proposal but are not a quantity of interest for inference.

        The sampled momentum variable sets out the proposal path for $\theta$ governed by the Hamiltonian dynamics (represented by a set of differential equations). These differential equations are solved using a discrete approximation algorithm, such as a leapfrog integrator, and is a function of the log posterior gradients for a given point $\theta$. The leapfrog integrator has a set of tuning parameters which determines the size and number of steps to be taken along the Hamiltonian path with the final step being the proposal value $(\theta^{\ast}, \phi^{\ast})$. Finally, the Metropolis Hastings accept/reject step is applied using the parameters at the beginning of the leapfrog process $(\theta^{t-1}, \phi^{t-1})$ and the proposal values $(\theta^{\ast}, \phi^{\ast})$. A formal description of the HMC algorithm is provided in Appendix C.

        \subsubsection{Implementation}
        The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler \citep{hoffman2014no}, allows for direct sampling of the specified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will also use the same priors as specified in the KSC's Gaussian mixture approximation so that HMC is sampling from the same model (although other priors could be chosen).

@@

        HMC on the other hand is a more general approach to applying MCMC. The implementation of HMC is through the Stan programming language which allows for the specfication other generative or Bayesian models, not just stochastic volatility. The implementation details of HMC is abstracted away from the user, who only needs to focus on specifying the model using the Stan syntax. Any hyperparameter tuning and initialisation is automated away. Furthermore, Stan and HMC allows for the specification of different priors, whereas KSC's bespoke algorithm is dependent on the use of conjugate priors.

@@

        The stochastic volatility model as described in section 2.1 is follows a centered parameterisation. This describes the central location of the latent state distribution which is centered on the mean of log volatility $\mu$. A model can be reparameterised in a way that improves MCMC performance. Reparameterisation applies a transformation to the model which may change the interpretation of the parameters or the way they are sampled but remains mathematically equivalent to the original model. The reparametiersations available for a model is dependent on the MCMC approach. Some reparamterisations are available only to specific sampling strategies and may improve or impair a MCMC performance.

        SBC is applied to reparameteirsied versions of the stochastic volatility model to see if there is any improvement in calibration and efficiency.

        \subsubsection{Reparameterised stochastic volatility for Hamiltonian Monte Carlo}
        The reparameterisation for Hamiltonian Monte Carlo changes how the latent states are sampled. Specifically, the states are first sampled from a standard normal distribution and are transformed to be samples from the state equation. As noted above, the centered parametersation samples the states centered around $\mu$.

@@

        Generate a the second state parameter conditional on prior simulation and first state variable.

@@

\section{Appendix C: Hamiltonian Monte Carlo algorithm description}
Let $\theta$ be a target parameter and $\phi$ be the auxiliary momemtum variable. The Hamiltonian is defined as:
